Handbook streamFind: backend

### 2022-08-04 handbooks/Handbook_backend_streamFind.Rmd
LFU12

```{r description-files, echo=FALSE}
df_files_desc <- data.frame(
  N. = seq_len(length(files)),
  file_name = basename(files),
  device = c(rep("Agilent Q-TOF", 24)),
  type = c(rep("centroided", 3), rep("profile", 3), rep("profile", 18)),
  mode = c(rep("MS/MS", 24)),
  polarity = c(rep("positive", 6),
               rep("negative", 3), rep("positive",3),
               rep("negative", 3), rep("positive",3),
               rep("negative", 3), rep("positive",3)),
  description = c(
    rep("Basic centroided MS data spiked with chemical and internal standards (CS and IS, respectively).", 3),
    rep("Basic profile MS data spiked with CS and IS.", 3),
    rep("Blank MS data spiked with IS.", 6),
    rep("Wastewater secondary effluent MS data spiked with IS.", 6),
    rep("Wastewater secondary effluent treated with ozonated strong water MS data spiked with IS.", 6)
  )
)

knitr::kable(df_files_desc, caption = "Files included in the streamFind package.") %>%
  column_spec(column = 2, width = "100px") %>%
  column_spec(column = 3, width = "60px") %>%
  kable_styling(font_size = 11.5, bootstrap_options = c("striped", "hover", "responsive"), fixed_thead = TRUE) %>%
  scroll_box(width = "100%", height = "600px")
```

@@

```{r show-table-chemicals, echo=FALSE}
db$ionization <- "positive"
db[, `:=`("mz_pos" = neutralMass + 1.0073, "mz_neg" = neutralMass - 1.0073 )]
db[grepl("neg", db$comment), ionization := "both"]
data.table::setnames(db, "mixIUTA", "tag")
db[tag %in% 1, tag := "CS"]

knitr::kable(db[, .(name, CAS, formula, neutralMass, rt, ionization, tag)], caption = "Chemical standards spiked.") %>%
  kable_styling(font_size = 11, bootstrap_options = c("striped", "hover", "responsive"), fixed_thead = TRUE) %>%
  scroll_box(width = "100%", height = "600px")
```

@@

#adding the profile analyses to msData
sp1 <- addAnalyses(sp1, analyses = aProfs)

@@

```{r metadata_structure}
#metadata of the fourth analyses in the msData
t(streamFind::metadata(sp1, analyses = 4))

#applied directly to the msAnalysis S4 class
t(streamFind::metadata(a1))
```

@@

```{r}

#as a named vector
meta_1 <- c("water", TRUE, "centroid")
names(meta_1) <- c("matrix", "spiked", "datatype")

@@

#optional, but when numbered list the order is taken as is
names(meta_vec_list) <- analyses(sp1)

@@

#as a data.frame (would also work with data.table)
meta_df_2 <- data.frame(
  analysis = analyses(sp1),
  matrix = rep("water_as_df", 6),
  spiked = rep(TRUE, 6),
  datatype = c(rep("centroid", 3), rep("profile", 3))
)

@@

As aforementioned, access to raw data is based on defining _m/z_ and retention time targets. The function `makeTargets` is used for comprehensively build targets. As shown below, multiple ways can be used to assemble targets but similarly returning a uniform `data.table` class object with a target list. Accessing and plotting S4 methods in streamFind use the `makeTargets` function to collect raw data from the analysis files. Therefore, the arguments `mz`, `rt`, `ppm`, `sec` and `id` are present in most S4 methods for accessing and plotting data as demonstrated in the following sections.

```{r make_targets}
#case 1
mz1 <- c(247.1651, 239.0628)
rt1 <- c(1075, 1157)
id1 <- c("target1", "target2")
targets1 <- makeTargets(mz = mz1, rt = rt1, ppm = 20, sec = 60, id = id1)

@@

#case 2
mz2 <- data.frame(mzmin = c(247.1626, 239.0604), mzmax = c(247.1676, 239.0652))
rt2 <- data.frame(rtmin = c(1045, 1127), rtmax = c(1105, 1187))
targets2 <- makeTargets(mz = mz2, rt = rt2)

@@

#case 3
mz3 <- data.frame(
  id = c("target1", "target2"),
  mz = c(247.1651, 239.0628),
  rt = c(1075, 1157)
)
targets3 <- makeTargets(mz = mz3, ppm = 20, sec = 60)

@@

#case 4
mz4 <- data.frame(
  id = c("target1", "target2"),
  mzmin = c(247.1626, 239.0604), mzmax = c(247.1676, 239.0652),
  rtmin = c(1045, 1127), rtmax = c(1105, 1187)
)
targets4 <- makeTargets(mz = mz4)

@@

@@

`r # TODO remove zeros when returning EIC table?`
`r # TODO add check for spectra/chroms is the object before querying the data?`

@@

```{r eics-method-2}
#EIC of targets from second analyses using msData class
eic2 <- EICs(sp1, analyses = 2, mz = targets4)
head(eic2, 6)
```

@@

```{r ploteics-interactive-2}
#the same plot as the first for data.table but interactive
plotEICs(a1, mz = targets4, interactive = TRUE)
```

@@

```{r xics-method}
#using the msData class object for second target with wide ppm range
xic1 <- XICs(sp1, analyses = 3:4, mz = mz1[2], rt = rt1[2], ppm = 50, sec = 30)
xic1
```

@@

```{r plotxics-method-3}
#for the msAnalysis, the second target with wide m/z and rt ranges
plotXICs(a1, mz = mz1[2], rt = rt1[2], ppm = 500, sec = 120)
```

@@

The S4 method applied above query the data from the raw data files as that is less disk memory intense. However, for certain cases the raw data might be needed entirely. Therefore, the data can be loaded and stored as a simple _data.table_ in the respective _msAnalysis_ (or other classes of the same level). Below we demonstrate how to load (via `loadRawData` S4 method) and access (via `spectra` or `chromatograms`) the raw spectra and chromatograms to/from an _msAnalaysis_ and _msData_ object. Both S4 methods (`spectra` or `chromatograms`) return a _data.table_ where further operations can be applied to subset the data, as demonstrated below.

```{r loadrawdata-msanalysis}
#load raw data to an msAnalysis object
a1 <- loadRawData(a1)

#example of accessing the six most intense fragments of the first target with msAnalysis
s_data <- spectra(a1)
s_data <- s_data[preMZ >= targets4$mzmin[1] & preMZ <= targets4$mzmax[1], ]
s_data <- s_data[rt >= targets4$rtmin[1] & rt <= targets4$rtmax[1], ]
s_data <- s_data[order(intensity, decreasing = TRUE), ]
s_data[1:6, ]
```

```{r loadrawdata-msdata}
#load raw data to the first three samples in an msData object.
sp1 <- loadRawData(sp1, analyses = 1:3)

#example of accessing the two most intense fragments of the first target with msData
s_data_2 <- spectra(sp1)
s_data_2 <- s_data_2[preMZ >= targets4$mzmin[1] & preMZ <= targets4$mzmax[1], ]
s_data_2 <- s_data_2[rt >= targets4$rtmin[1] & rt <= targets4$rtmax[1], ]
s_data_2 <- s_data_2[order(intensity, decreasing = TRUE), ]
s_data_2[1:4, ]
```

@@

```{r compatible-vendor-formats, echo=FALSE}
vfor <- streamFind::compatibleFileFormatsForConversion()
knitr::kable(vfor, caption = "Vendor formats possible to be converted to mzML/mzXML.") %>%
  kable_styling(font_size = 11, bootstrap_options = c("striped", "hover", "responsive"), fixed_thead = TRUE) %>%
  scroll_box(width = "100%", height = "600px")
```

@@

An essential aspect of environmental analysis is to operate with sample/analysis replicates. The replicates in _msData_ or _msAnalysis_ can be obtained by the `replicates` S4 method as shown below. The setter (assignment) for the replicate names is demonstrated below with the method `replicates<-` that takes a vector of character strings with the same length as the number of analyses in a _msData_ object or length one for a single analysis in _msAnalysis_. Alternatively, a table with a replicate character column or a character vector can be given in the files or replicates arguments, respectively.

```{r assign-replicates}
#getter for replicate names in msData
replicates(sp2)

#setter for replicate names in msData
replicates(sp2) <- c(rep("blank", 3),rep("influent", 3))  
replicates(sp2)

#setter for replicate name in msAnalysis
a2 <- getAnalyses(sp2, 4)
replicates(a2) <- "wastewater"

#getter for replicate name in msAnalysis
replicates(a2)
```

### Assign blanks

@@

```{r assign-blanks}
#getter for blank replicates in each sample in a msData object
blanks(sp2)

#setter for the blank analysis replicate of msData object
blanks(sp2) <- rep("blank", 6)
blanks(sp2)

#getter for msAnalysis
blanks(a2)
```

@@

```{r other-s4methods}
#getter for analysis names in msdata
analyses(sp2)

#getter for analysis name in msAnalysis
analyses(a2)

@@

param <- xcms::CentWaveParam(
  ppm = 10, peakwidth = c(5, 60),
  snthresh = 5, prefilter = c(4, 800),
  mzCenterFun = "mean", integrate = 2,
  mzdiff = -0.0001, fitgauss = TRUE,
  noise = 250, verboseColumns = TRUE,
  firstBaselineCheck = FALSE,
  extendLengthMSW = TRUE
)

@@

# the algorithm of the influent samples is different
sapply(analyses(sp2_pp_3), function(x) getAlgorithm(getParameters(sp2_pp_3, where = "analyses", analyses = x)[[1]]))
```

@@

```{r peakpicking-multiple-show}
# as the peak parameters with "openms" as algorithm were less stringent, the influent samples have more peaks then when "xcms3" is used 
sp2_pp_3
```

@@

```{r inspecting-peaks-msdata-table}
#access to peaks in msData based on built targets as used for extracting EICs
pks_1 <- peaks(sp2_pp[c(1, 4)], mz = mz1[1], rt = rt1[1], ppm = 10, sec = 30)

knitr::kable(pks_1, caption = "Peaks data.table from msData using the peaks S4 method.") %>%
  kable_styling(font_size = 11, bootstrap_options = c("striped", "hover", "responsive"), fixed_thead = TRUE) %>%
  scroll_box(width = "100%", height = "600px")
```

@@

@@

# setInfo(msd)
# path(msd)
# files(msd)
# analyses(msd)
# replicates(msd)
# blanks(msd)
# polarities(msd)
# analysisInfo(msd)
# 
# replicates(msd) <- c("s1", "s2")
# blanks(msd) <- c(NA_character_, "s2")
# 
# validObject(msd)
# 
# files(msa_1)
# analyses(msa_1)
# replicates(msa_1)
# blanks(msa_1)
# polarities(msa_1)
# 
# replicates(msa_1) <- "s1"
# blanks(msa_1) <- "s2"

# object <- msd
# 
# loadMS1 <- function(object, analysis = NULL, what = "MS1") {
#   
#   
#   test <- RaMS::grabMSdata(files(object)[1], grab_what = "MS1", rtrange = c(min(rt_02), max(rt_02)))
#   
#   ms2 <- test$MS2
#   
#   ms2[premz > 273 & premz < 274, ]
#   
#   View(ms2)
# 
# }

>>>>

```{r description-files, echo=FALSE}
df_files_desc <- data.frame(
  N. = seq_len(length(files)),
  file_type = rep("MS", 6 + 18),
  file_name = basename(files),
  device = c(rep("Agilent Q-TOF", 24)),
  data_type = c(rep("centroid", 3), rep("profile", 3), rep("centroid", 18)),
  ac_mode = c(rep("MS/MS", 24)),
  polarity = c(rep("positive", 6),
               rep("negative", 3), rep("positive",3),
               rep("negative", 3), rep("positive",3),
               rep("negative", 3), rep("positive",3)),
  description = c(
    rep("Basic centroided MS data spiked with chemical and internal standards (CS and IS, respectively).", 3),
    rep("Basic profile MS data spiked with CS and IS.", 3),
    rep("Blank MS data spiked with IS.", 6),
    rep("Wastewater secondary effluent MS data spiked with IS.", 6),
    rep("Wastewater secondary effluent treated with ozonated strong water MS data spiked with IS.", 6)
  )
)

knitr::kable(df_files_desc, caption = "Files included in the streamFind package.") %>%
  column_spec(column = 3, width = "100px") %>%
  column_spec(column = 4, width = "60px") %>%
  kable_styling(font_size = 11.5, bootstrap_options = c("striped", "hover", "responsive"), fixed_thead = TRUE) %>%
  scroll_box(width = "100%", height = "600px")
```

@@

```{r show-table-chemicals, echo=FALSE}
db$ionization <- "positive"
db[, `:=`("mz_pos" = neutralMass + 1.0073, "mz_neg" = neutralMass - 1.0073 )]
db[grepl("neg", db$comment), ionization := "both"]
data.table::setnames(db, "mixIUTA", "tag")
db[tag %in% 1, tag := "CS"]
db[, in_file := "1-6"]
db[tag %in% "IS", in_file := "1-24"]

knitr::kable(db[, .(name, CAS, formula, neutralMass, rt, ionization, tag, in_file)], caption = "Chemical standards spiked.") %>%
  kable_styling(font_size = 11, bootstrap_options = c("striped", "hover", "responsive"), fixed_thead = TRUE) %>%
  scroll_box(width = "100%", height = "600px")
```

@@

#adding the profile analyses to msData
sp1 <- addAnalyses(sp1, analysisList = aProfs)

@@

```{r metadata_structure}
#metadata of the fourth analyses in the msData
t(metadata(sp1, analyses = 4))

#applied directly to the msAnalysis S4 class
t(metadata(a1))
```

@@

```{r}
#as a named vector
meta_1 <- c("water", TRUE, "centroid")
names(meta_1) <- c("matrix", "spiked", "datatype")

@@

#optional, but when numbered list the order is taken as is
names(meta_vec_list) <- analysisNames(sp1)

@@

#as a data.frame (would also work with data.table)
meta_df_2 <- data.frame(
  analysis = analysisNames(sp1),
  matrix = rep("water_as_df", 6),
  spiked = rep(TRUE, 6),
  datatype = c(rep("centroid", 3), rep("profile", 3))
)

@@

As aforementioned, access to raw data is based on defining _m/z_ and retention time targets. The function `makeTargets` is used for comprehensively build targets. As shown below for Carbamazepin d10 and Diuron d6, multiple ways can be used to assemble targets but similarly returning a uniform `data.table` class object with a target list. Accessing and plotting S4 methods in streamFind use the `makeTargets` function to collect raw data from the analysis files. Therefore, the arguments `mz`, `rt`, `ppm`, `sec` and `id` are present in most S4 methods for accessing and plotting data as demonstrated in the following sections.

```{r carb-diu}
carb <- db[name %in% "Carbamazepin d10", .(name, neutralMass, rt)]
carb$mz <- carb$neutralMass + 1.0073
carb

diu <- db[name %in% "Diuron d6", .(name, neutralMass, rt)]
diu$mz <- diu$neutralMass + 1.0073
diu
```

```{r make_targets}
#case 1
mz1 <- c(247.1650, 239.0620)
rt1 <- c(1075, 1157)
id1 <- c("target1", "target2")
targets1 <- makeTargets(mz = mz1, rt = rt1, ppm = 20, sec = 30, id = id1)

@@

#case 2
mz2 <- data.frame(mzmin = c(247.1601, 239.0572), mzmax = c(247.1699, 239.0668))
rt2 <- data.frame(rtmin = c(1045, 1127), rtmax = c(1105, 1187))
targets2 <- makeTargets(mz = mz2, rt = rt2)

@@

#case 3
mz3 <- data.frame(
  id = c("target1", "target2"),
  mz = c(247.1650, 239.0620),
  rt = c(1075, 1157)
)
targets3 <- makeTargets(mz = mz3, ppm = 20, sec = 30)

@@

#case 4
mz4 <- data.frame(
  id = c("target1", "target2"),
  mzmin = c(247.1601, 239.0572), mzmax = c(247.1699, 239.0668),
  rtmin = c(1045, 1127), rtmax = c(1105, 1187)
)
targets4 <- makeTargets(mz = mz4)

@@

```{r test-targets}
#test all equal target 1
t1 <- rbind(targets1[1, ], targets2[1,], targets3[1, ], targets4[1, ])
all(round(apply(t1[, 2:7], 2, sd), digits = 4) == 0)

#test all equal target 2
t2 <- rbind(targets1[2, ], targets2[2,], targets3[2, ], targets4[2, ])
all(round(apply(t2[, 2:7], 2, sd), digits = 4) == 0)
```

@@

@@

```{r eics-method-2}
#EIC of targets from analyses 1 to 3 using msData class
eic2 <- EICs(sp1, analyses = 1:3, mz = targets4)
head(eic2, 6)
```

@@

```{r ploteics-interactive-2}
#using msAnalysis
plotEICs(a1, mz = targets4, interactive = TRUE)
```

```{r ploteics-interactive-3}
#using msData centroid (file 3) vs profile (file 4); note that the higher intensity for profile data is due to the sum of the traces within the same spectrum (i.e., with the same retention time)
plotEICs(sp1, analyses = 3:4, mz = targets3, colorBy = "replicates", interactive = TRUE)
```

@@

```{r xics-method}
#using the msData class object for second target with wide ppm range
xic1 <- XICs(sp1, analyses = 3:4, mz = mz1[2], rt = rt1[2], ppm = 500, sec = 30)
xic1
```

@@

```{r plotxics-method-3}
#for the msAnalysis, the second target with wide m/z and rt ranges
plotXICs(a1, mz = mz1[2], rt = rt1[2], ppm = 50, sec = 120)
```

@@

The S4 methods applied above parse the data from the raw data files which is less disk memory intense but slower. Alternatively, the raw data can be loaded to (each) _msAnalysis_  and from their is faster available. However, when loaded, the raw data has a higher memory footprint.
Therefore, the data can be optionally loaded and stored as a simple _data.table_ in the respective _msAnalysis_ (other classes of the same level might have the raw data directly loaded if the data has a lower memory footprint). Below we demonstrate how to load (via `loadRawData` S4 method) and access (via `spectra` or `chromatograms`) the raw spectra and chromatograms from an _msAnalaysis_ and _msData_ object. Both S4 methods (`spectra` or `chromatograms`) return a _data.table_ where further operations can be applied to subset the data, as demonstrated below. The standard _data.table_ operations can be applied, see documentation in https://cran.r-project.org/web/packages/data.table/ for more information and code examples/instructions. A cheat sheet for _data.table_ operations can be downloaded in https://raw.githubusercontent.com/rstudio/cheatsheets/main/datatable.pdf. In streamFind, the _data.table_ is always the preferred format for table objects.

```{r loadrawdata-msanalysis}
#load raw data to an msAnalysis object, while setting a minimum intensity value for both MS1 and MS2 traces
a1 <- loadRawData(a1, minIntensityMS1 = 200, minIntensityMS2 = 100)
```

```{r loadrawdata-msanalysis-show}
a1
```

```{r loadrawdata-msdata}
#load raw data to the first three samples in an msData object.
sp1 <- loadRawData(sp1, analyses = 1:3)
```

```{r loadrawdata-msdata-show}
sp1
```

## Spectra

```{r spectra-msanalysis}
#example of accessing the six most intense fragments of the first target with msAnalysis
s_data <- spectra(a1)
s_data <- s_data[preMZ >= targets4$mzmin[1] & preMZ <= targets4$mzmax[1], ]
s_data <- s_data[rt >= targets4$rtmin[1] & rt <= targets4$rtmax[1], ]
s_data <- s_data[order(intensity, decreasing = TRUE), ]
s_data[1:6, ]
```

```{r spectra-msdata}
#example of accessing the two most intense fragments of the first target with msData
s_data_2 <- spectra(sp1)
s_data_2 <- s_data_2[preMZ >= targets4$mzmin[1] & preMZ <= targets4$mzmax[1], ]
s_data_2 <- s_data_2[rt >= targets4$rtmin[1] & rt <= targets4$rtmax[1], ]
s_data_2 <- s_data_2[order(intensity, decreasing = TRUE), ]
s_data_2[1:4, ]
```

## Chromatograms

@@

```{r compatible-vendor-formats, echo=FALSE}
vfor <- streamFind::compatibleFileFormatsForConversion()
knitr::kable(vfor, caption = "Vendor formats possible to be converted to mzML/mzXML.") %>%
  kable_styling(font_size = 11, bootstrap_options = c("striped", "hover", "responsive"), fixed_thead = TRUE) %>%
  scroll_box(width = "100%", height = "300px")
```

@@

An essential aspect of environmental analysis is to operate with sample/analysis replicates. The replicates in _msData_ or _msAnalysis_ can be obtained by the `replicateNames` S4 method as shown below. The setter (assignment) for the replicate names is demonstrated below with the method `replicateNames<-` that takes a vector of character strings with the same length as the number of analyses in a _msData_ object or length one for a single analysis in _msAnalysis_. Alternatively, a table with a replicate character column or a character vector can be given in the files or replicates arguments, respectively.

```{r assign-replicatenames}
#getter for replicate names in msData
replicateNames(sp2)

#setter for replicate names in msData
replicateNames(sp2) <- c(rep("blank", 3),rep("influent", 3))  
replicateNames(sp2)

#setter for replicate name in msAnalysis
a2 <- getAnalyses(sp2, 4)
replicateNames(a2) <- "wastewater"

#getter for replicate name in msAnalysis
replicateNames(a2)
```

### Assign blankReplicateNames

@@

```{r assign-blankreplicatenames}
#getter for blank replicates in each sample in a msData object
blankReplicateNames(sp2)

#setter for the blank analysis replicate of msData object
blankReplicateNames(sp2) <- rep("blank", 6)
blankReplicateNames(sp2)

#getter for msAnalysis
blankReplicateNames(a2)
```

@@

```{r other-s4methods}
#getter for analysis names in msdata
analysisNames(sp2)

#getter for analysis name in msAnalysis
analysisNames(a2)

@@

param <- xcms::CentWaveParam(
  ppm = 12, peakwidth = c(5, 40),
  snthresh = 5, prefilter = c(4, 800),
  mzCenterFun = "mean", integrate = 2,
  mzdiff = -0.0001, fitgauss = TRUE,
  noise = 250, verboseColumns = TRUE,
  firstBaselineCheck = FALSE,
  extendLengthMSW = TRUE
)

@@

# the algorithm of the influent samples is different
sapply(analysisNames(sp2_pp_3), function(x) getAlgorithm(getParameters(sp2_pp_3, where = "analyses", analyses = x)[[1]]))
```

@@

```{r peakpicking-multiple-show}
# as the peak parameters with "openms" as algorithm were more stringent, the influent analyses have less peaks when compared to "xcms3" 
sp2_pp_3
```

@@

```{r inspecting-peaks-msdata-table}
#access to peaks in msData based on built targets as used for extracting EICs
peaks(sp2_pp[c(1, 4)], mz = mz1[1], rt = rt1[1], ppm = 10, sec = 30)
```

@@

# knitr::kable(pks_1, caption = "Peaks data.table from msData using the peaks S4 method.") %>%
#   kable_styling(font_size = 11, bootstrap_options = c("striped", "hover", "responsive"), fixed_thead = TRUE) %>%
#   scroll_box(width = "100%", height = "600px")

@@

# setInfo(msd)
# path(msd)
# filePaths(msd)
# analysisNames(msd)
# replicateNames(msd)
# blankReplicateNames(msd)
# polarities(msd)
# analysisInfo(msd)
# 
# replicateNames(msd) <- c("s1", "s2")
# blankReplicateNames(msd) <- c(NA_character_, "s2")
# 
# validObject(msd)
# 
# filePaths(msa_1)
# analysisNames(msa_1)
# replicateNames(msa_1)
# blankReplicateNames(msa_1)
# polarities(msa_1)
# 
# replicateNames(msa_1) <- "s1"
# blankReplicateNames(msa_1) <- "s2"

# object <- msd
# 
# loadMS1 <- function(object, analysis = NULL, what = "MS1") {
#   
#   
#   test <- RaMS::grabMSdata(filePaths(object)[1], grab_what = "MS1", rtrange = c(min(rt_02), max(rt_02)))
#   
#   ms2 <- test$MS2
#   
#   ms2[premz > 273 & premz < 274, ]
#   
#   View(ms2)
# 
# }
